---
title: "Logistic Regression: Heart Disease Prediction"
author: "Osman Omer Mustafa"
date: "today"
format:
  html:
    toc: true
    code-fold: true
    theme: cosmo
    toc-location: left
execute:
  warning: false
  message: false
---

## Introduction

**Purpose:** This analysis demonstrates logistic regression for binary classification 
in a medical context. The goal is to predict heart disease presence (yes/no) using 
clinical and diagnostic variables.

**Educational Context:** This project is created for learning purposes to understand:

- Binary classification modeling
- Odds ratios interpretation
- Model evaluation metrics (confusion matrix, ROC/AUC)
- Clinical vs comprehensive diagnostic models
- Limitations of observational data

**Important Note:** This analysis uses publicly available data and is intended solely 
for educational demonstration of statistical techniques. Results should not be used 
for actual clinical decision-making without proper validation.

## Load required libraries
```{r setup}
library(tidyverse)
library(broom)
library(pROC)
library(caret)
```

## Read data
```{r}
data <- read_csv("heart.csv")
```

## Basic checks
```{r}
glimpse(data)
summary(data)
sum(is.na(data))
```

## Outcome distribution
```{r}
table(data$target)
prop.table(table(data$target))
```

```{r echo=FALSE}
outcome_counts <- table(data$target)
outcome_props <- prop.table(outcome_counts)
n_no <- outcome_counts[1]
n_yes <- outcome_counts[2]
prop_no <- outcome_props[1]
prop_yes <- outcome_props[2]
```

The dataset contains `r n_no` individuals without heart disease (`r round(prop_no*100, 1)`%) 
and `r n_yes` with heart disease (`r round(prop_yes*100, 1)`%). This represents a 
**balanced dataset**, which is ideal for logistic regression modeling. Balanced classes 
ensure the model doesn't develop bias toward the majority class.

## Ensure proper data types
```{r}
data <- data %>%
  mutate(
    target = as.factor(target),
    sex = as.factor(sex),
    cp = as.factor(cp),
    fbs = as.factor(fbs),
    restecg = as.factor(restecg),
    exang = as.factor(exang),
    slope = as.factor(slope),
    ca = as.factor(ca),
    thal = as.factor(thal)
  )

glimpse(data)
```

Categorical variables have been converted to factors to ensure proper treatment in 
logistic regression. R will automatically create dummy variables for these predictors.

## Exploratory Data Analysis

### Outcome by sex
```{r}
outcome_sex <- data %>%
  group_by(sex, target) %>%
  summarise(n = n(), .groups = "drop") %>%
  group_by(sex) %>%
  mutate(proportion = n / sum(n))

outcome_sex %>% knitr::kable(caption = "Heart Disease by Sex", digits = 3)
```

**Observation:** `r if(outcome_sex$proportion[outcome_sex$sex == "0" & outcome_sex$target == "1"] > outcome_sex$proportion[outcome_sex$sex == "1" & outcome_sex$target == "1"]) {"Females show higher disease prevalence than males in this dataset, which contradicts typical population patterns. This likely reflects referral bias or specific sampling characteristics."} else {"Males show higher disease prevalence, consistent with typical cardiovascular disease patterns."}`

### Age distribution by outcome
```{r}
ggplot(data, aes(x = target, y = age, fill = target)) +
  geom_boxplot() +
  labs(
    x = "Heart Disease",
    y = "Age (Years)",
    title = "Age Distribution by Heart Disease Status"
  ) +
  scale_x_discrete(labels = c("No", "Yes")) +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5), legend.position = "none")
```

Visual comparison of age distributions between groups. Note any unexpected patterns 
that may indicate sampling bias.

### Max heart rate by outcome
```{r}
ggplot(data, aes(x = target, y = thalach, fill = target)) +
  geom_boxplot() +
  labs(
    x = "Heart Disease",
    y = "Max Heart Rate",
    title = "Max Heart Rate by Heart Disease Status"
  ) +
  scale_x_discrete(labels = c("No", "Yes")) +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5), legend.position = "none")
```

Maximum heart rate achieved during exercise testing often differs between healthy 
individuals and those with cardiovascular disease.

## Logistic Regression Models

### Understanding Binary Outcomes

Unlike linear regression (which predicts continuous values), logistic regression 
predicts **probabilities** for binary outcomes (0 or 1).

**Key concept:** The model uses the **logistic function** to ensure predicted 
probabilities stay between 0 and 1:

```
P(disease = 1) = 1 / (1 + e^(-(β₀ + β₁×age + β₂×sex + ...)))
```

**Coefficients (β) are interpreted as log odds ratios.** When exponentiated 
(`exponentiate = TRUE`), they become **odds ratios (OR)**.

### Model 1: Age only
```{r}
model_age <- glm(target ~ age, data = data, family = binomial)

tidy_age <- tidy(model_age, conf.int = TRUE, exponentiate = TRUE)
tidy_age %>% knitr::kable(caption = "Age Model Coefficients (Odds Ratios)", digits = 3)
```

```{r echo=FALSE}
age_or <- tidy_age$estimate[tidy_age$term == "age"]
age_pval <- tidy_age$p.value[tidy_age$term == "age"]
age_sig <- age_pval < 0.05
```

**Interpretation:** The odds ratio for age is `r round(age_or, 3)`. 
`r if(age_or < 1) {paste0("This suggests that for each additional year of age, the odds of heart disease decrease by ", round((1-age_or)*100, 1), "%. This contradicts typical medical expectations and likely reflects sampling bias in this dataset.")} else {paste0("This indicates that for each additional year of age, the odds of heart disease increase by ", round((age_or-1)*100, 1), "%.")}`

### Model 2: Clinical model
```{r}
model_clinical <- glm(target ~ age + sex + cp + trestbps + chol + fbs + thalach, 
                      data = data, family = binomial)

tidy_clinical <- tidy(model_clinical, conf.int = TRUE, exponentiate = TRUE)
tidy_clinical %>% knitr::kable(caption = "Clinical Model Coefficients (Odds Ratios)", digits = 3)
```

```{r echo=FALSE}
glance_clinical <- glance(model_clinical)
dev_clinical <- glance_clinical$deviance
```

**Clinical model includes:** Basic clinical measurements obtainable in routine practice 
without specialized testing (age, sex, chest pain type, blood pressure, cholesterol, 
fasting blood sugar, max heart rate).

Model deviance: `r round(dev_clinical, 1)` (lower is better).

### Model 3: Full model
```{r}
model_full <- glm(target ~ age + sex + cp + trestbps + chol + fbs + restecg +
                    thalach + exang + oldpeak + slope + ca + thal,
                  data = data, family = binomial)

tidy_full <- tidy(model_full, conf.int = TRUE, exponentiate = TRUE)
tidy_full %>% knitr::kable(caption = "Full Model Coefficients (Odds Ratios)", digits = 3)
```



```{r echo=FALSE}
glance_full <- glance(model_full)
dev_full <- glance_full$deviance
```

**Full model includes:** All available diagnostic variables including ECG findings, 
exercise test results, and angiography data.

Model deviance: `r round(dev_full, 1)` (substantially lower than clinical model).

## Model Comparison

### Deviance comparison
```{r}
deviance_comparison <- data.frame(
  model = c("Clinical", "Full"),
  deviance = c(dev_clinical, dev_full),
  improvement = c(1420 - dev_clinical, 1420 - dev_full)
)
deviance_comparison %>% knitr::kable(caption = "Model Deviance Comparison", digits = 1)
```

**Null deviance** (intercept-only model) = 1420. Lower deviance indicates better fit.

Clinical model reduced deviance by `r round(1420 - dev_clinical, 1)` units 
(`r round((1420 - dev_clinical)/1420 * 100, 1)`% improvement).

Full model reduced deviance by `r round(1420 - dev_full, 1)` units 
(`r round((1420 - dev_full)/1420 * 100, 1)`% improvement).

Full model provides an additional `r round(dev_clinical - dev_full, 1)` unit improvement 
over the clinical model (`r round((dev_clinical - dev_full)/dev_clinical * 100, 1)`%).

### AIC/BIC comparison
```{r}
aic_comparison <- data.frame(
  model = c("Clinical", "Full"),
  AIC = c(AIC(model_clinical), AIC(model_full)),
  BIC = c(BIC(model_clinical), BIC(model_full))
)
aic_comparison %>% knitr::kable(caption = "AIC/BIC Comparison", digits = 1)
```

Both AIC and BIC favor the full model (lower values = better). The substantial 
improvement justifies the added complexity.

## Model Predictions

```{r}
data$pred_prob_clinical <- predict(model_clinical, type = "response")
data$pred_prob_full <- predict(model_full, type = "response")

data$pred_class_clinical <- ifelse(data$pred_prob_clinical > 0.5, "1", "0")
data$pred_class_full <- ifelse(data$pred_prob_full > 0.5, "1", "0")
```

Predicted probabilities converted to classifications using threshold = 0.5. 
Alternative thresholds may be explored for different clinical contexts 
(e.g., lowering threshold increases sensitivity at cost of specificity).

## Model Evaluation

### Confusion Matrix: Clinical model
```{r}
conf_matrix_clinical <- confusionMatrix(
  as.factor(data$pred_class_clinical),
  data$target,
  positive = "1"
)
conf_matrix_clinical
```

```{r echo=FALSE}
acc_clin <- conf_matrix_clinical$overall["Accuracy"]
sens_clin <- conf_matrix_clinical$byClass["Sensitivity"]
spec_clin <- conf_matrix_clinical$byClass["Specificity"]
ppv_clin <- conf_matrix_clinical$byClass["Pos Pred Value"]
npv_clin <- conf_matrix_clinical$byClass["Neg Pred Value"]
```

**Clinical Model Performance:**

- **Accuracy:** `r round(acc_clin, 3)` (`r round(acc_clin*100, 1)`%)
- **Sensitivity:** `r round(sens_clin, 3)` (`r round(sens_clin*100, 1)`%) - detects `r round(sens_clin*100, 1)`% of actual disease cases
- **Specificity:** `r round(spec_clin, 3)` (`r round(spec_clin*100, 1)`%) - correctly identifies `r round(spec_clin*100, 1)`% of healthy individuals
- **PPV:** `r round(ppv_clin, 3)` - when model predicts disease, it's correct `r round(ppv_clin*100, 1)`% of the time
- **NPV:** `r round(npv_clin, 3)` - when model predicts no disease, it's correct `r round(npv_clin*100, 1)`% of the time

### Confusion Matrix: Full model
```{r}
conf_matrix_full <- confusionMatrix(
  as.factor(data$pred_class_full),
  data$target,
  positive = "1"
)
conf_matrix_full
```

```{r echo=FALSE}
acc_full <- conf_matrix_full$overall["Accuracy"]
sens_full <- conf_matrix_full$byClass["Sensitivity"]
spec_full <- conf_matrix_full$byClass["Specificity"]
ppv_full <- conf_matrix_full$byClass["Pos Pred Value"]
npv_full <- conf_matrix_full$byClass["Neg Pred Value"]
```

**Full Model Performance:**

- **Accuracy:** `r round(acc_full, 3)` (`r round(acc_full*100, 1)`%)
- **Sensitivity:** `r round(sens_full, 3)` (`r round(sens_full*100, 1)`%) 
- **Specificity:** `r round(spec_full, 3)` (`r round(spec_full*100, 1)`%)
- **PPV:** `r round(ppv_full, 3)`
- **NPV:** `r round(npv_full, 3)`

### Performance comparison
```{r}
metrics_clinical <- data.frame(
  model = "Clinical",
  accuracy = acc_clin,
  sensitivity = sens_clin,
  specificity = spec_clin,
  ppv = ppv_clin,
  npv = npv_clin
)

metrics_full <- data.frame(
  model = "Full",
  accuracy = acc_full,
  sensitivity = sens_full,
  specificity = spec_full,
  ppv = ppv_full,
  npv = npv_full
)

metrics_comparison <- bind_rows(metrics_clinical, metrics_full)
metrics_comparison %>% knitr::kable(caption = "Model Performance Comparison", digits = 3)
```

The full model shows improvement across all metrics:

- Accuracy: +`r round((acc_full - acc_clin)*100, 1)`%
- Sensitivity: +`r round((sens_full - sens_clin)*100, 1)`% (fewer missed diagnoses)
- Specificity: +`r round((spec_full - spec_clin)*100, 1)`% (fewer false alarms)
- NPV: +`r round((npv_full - npv_clin)*100, 1)`% (greater confidence in negative predictions)

## ROC Curve and AUC

```{r}
roc_clinical <- roc(data$target, data$pred_prob_clinical)
auc_clinical <- auc(roc_clinical)

roc_full <- roc(data$target, data$pred_prob_full)
auc_full <- auc(roc_full)
```

```{r}
plot(roc_clinical, col = "blue", main = "ROC Curves Comparison", 
     lwd = 2, legacy.axes = TRUE)
plot(roc_full, col = "red", add = TRUE, lwd = 2)
legend("bottomright", 
       legend = c(paste0("Clinical (AUC = ", round(auc_clinical, 3), ")"),
                  paste0("Full (AUC = ", round(auc_full, 3), ")")),
       col = c("blue", "red"), lwd = 2)
```

### Understanding ROC and AUC

**ROC Curve:** Plots sensitivity (y-axis) vs. 1-specificity (x-axis) across all 
possible classification thresholds. A perfect classifier reaches the top-left corner 
(100% sensitivity, 100% specificity). The diagonal line represents random guessing (AUC = 0.5).

**AUC (Area Under the Curve):** Measures overall model performance. Interpretation:

- 0.5 = Random guessing (no discriminative ability)
- 0.6-0.7 = Poor
- 0.7-0.8 = Fair
- 0.8-0.9 = Good
- 0.9-1.0 = Excellent

**Clinical model AUC = `r round(auc_clinical, 3)`:** Good discriminative ability, 
suitable for screening applications.

**Full model AUC = `r round(auc_full, 3)`:** Excellent discriminative ability, 
suitable for diagnostic applications. The `r round((auc_full - auc_clinical)*100, 1)`% 
improvement over the clinical model is substantial.

## Visualization: Predicted Probabilities

```{r}
ggplot(data, aes(x = pred_prob_clinical, fill = target)) +
  geom_histogram(bins = 30, alpha = 0.7, position = "identity") +
  labs(
    x = "Predicted Probability",
    y = "Frequency",
    title = "Clinical Model: Predicted Probabilities by Actual Outcome",
    fill = "Heart Disease"
  ) +
  scale_fill_discrete(labels = c("No", "Yes")) +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5))
```

Good separation between the two outcome groups indicates strong predictive performance. 
Overlap represents cases where the model struggles to classify correctly.

## Limitations and Considerations

### Dataset Limitations

**1. Sampling Bias:**
This dataset shows unusual patterns (e.g., younger individuals with higher disease 
prevalence, females with higher prevalence than males) that contradict typical 
population epidemiology. This likely reflects:

- Referral bias (symptomatic young patients more likely to be tested)
- Hospital-based sampling (not representative of general population)
- Survivor bias (older individuals without severe disease)

**2. Generalizability:**
Results may not apply to:

- General population screening
- Different geographic/demographic groups
- Prevention-focused rather than diagnostic contexts

**3. Sample Size:**
While adequate for this analysis (n = `r nrow(data)`), larger datasets would enable:

- More robust validation (train/test splits)
- Detection of rare but important patterns
- Subgroup analyses

### Modeling Limitations

**1. No Validation Set:**
This analysis uses the entire dataset for both training and evaluation, which may 
overestimate performance. Proper validation requires:

- Train/test split (e.g., 80/20)
- Cross-validation
- External validation on independent data

**2. Class Imbalance (Not an Issue Here):**
This dataset is well-balanced (`r round(prop_yes*100, 1)`% diseased). 
Imbalanced datasets require special handling (oversampling, SMOTE, cost-sensitive learning).

**3. Threshold Selection:**
We used the standard 0.5 threshold. Clinical applications may require:

- Lower thresholds for screening (prioritize sensitivity)
- Higher thresholds for confirmatory testing (prioritize specificity)
- Cost-benefit analysis to determine optimal cutoff

**4. Missing Interaction Terms:**
This analysis doesn't explore interactions (e.g., age × smoking, BMI × diabetes). 
These could improve predictions but increase model complexity.

### Clinical Interpretation Cautions

**1. Correlation ≠ Causation:**
These models identify predictive associations, not causal relationships. Observational 
data cannot prove that changing a risk factor will alter outcomes.

**2. Odds Ratios vs. Probabilities:**
Odds ratios (ORs) are commonly misinterpreted:

- OR = 2 does NOT mean "twice the probability"
- OR = 2 means "twice the odds" (probability/(1-probability))
- For rare outcomes, OR ≈ relative risk
- For common outcomes (like in this dataset), OR ≠ relative risk

**3. Unexpected Coefficients:**
Several variables show counter-intuitive effects (age, blood pressure) due to 
confounding and sampling bias. These should not inform clinical practice.

### Educational Purpose

**This analysis demonstrates:**

- ✅ Logistic regression methodology
- ✅ Model comparison techniques
- ✅ Evaluation metrics interpretation
- ✅ ROC/AUC analysis

**This analysis does NOT provide:**

- ❌ Clinically validated diagnostic tools
- ❌ Guidance for patient care
- ❌ Generalizable epidemiological insights

**For actual clinical applications, models require:**

- Prospective validation studies
- Regulatory approval
- Integration with clinical workflows
- Ongoing monitoring and recalibration

## Conclusion

This analysis demonstrates logistic regression for binary classification in a 
cardiovascular disease prediction context. Key findings include:

**Model Performance:**

The full diagnostic model (AUC = `r round(auc_full, 3)`) substantially outperformed 
the clinical model (AUC = `r round(auc_clinical, 3)`), achieving excellent 
discriminative ability. Sensitivity improved from `r round(sens_clin*100, 1)`% to 
`r round(sens_full*100, 1)`%, reducing missed diagnoses by approximately 
`r round((1 - (1-sens_full)/(1-sens_clin))*100, 0)`%.

**Practical Implications:**

While the full model's superior performance is clear, implementation requires:

- Access to specialized diagnostic tests (ECG, exercise testing, angiography)
- Higher cost and time investment
- Trained personnel for test administration

The clinical model, though less accurate, uses readily available measurements and 
may suffice for initial screening in resource-limited settings.

**Key Limitations:**

The dataset exhibits unusual epidemiological patterns suggesting referral and 
sampling biases. Results should be interpreted as educational demonstrations of 
statistical methods rather than clinically applicable findings. Proper validation 
on independent, representative samples is essential before any clinical application.

**Methodological Learning:**

This analysis successfully demonstrates:

- Logistic regression for binary outcomes
- Model comparison using deviance, AIC/BIC
- Comprehensive evaluation with confusion matrices
- ROC curve analysis and AUC interpretation
- Importance of considering limitations

**Future Directions:**

To improve upon this analysis:

- Implement train/test validation
- Explore interaction terms
- Test alternative thresholds for different clinical contexts
- Validate on external datasets
- Compare with machine learning approaches (random forests, neural networks)

**Final Note:**

This project serves its educational purpose: demonstrating how logistic regression 
can model binary health outcomes and how to rigorously evaluate such models. 
The techniques learned here are applicable to any binary classification problem 
in medical research, epidemiology, or data science.
